---
title: "evalid: Methods for Improving External Validity of Randomized Experiments"
output: 
  md_document:
    variant: markdown_github
---

# evalid: Methods for Improving External Validity of Randomized Experiments

**Description:** 

R package `evalid` provides two key functions. (1) `tpate()`: three classes of estimators for the target population average treatment effect (T-PATE) to conduct effect-generalization. (2) `pct()`: a partial conjunction test to conduct a sign-generalization test. Details of the methods are described in Egami and Hartman (2022+).

**Authors:** 

- [Naoki Egami](https://naokiegami.com/)
- [Erin Hartman](https://erinhartman.com/)

**Reference:** 

- Egami and Hartman. (2022+). [Elements of External Validity: Framework, Design, and Analysis.](https://naokiegami.com/paper/external_full.pdf) Conditionally accepted at *American Political Science Review*

## Installation Instructions
You can install the most recent development version using the `devtools` package. First you have to install `devtools` using the following code. Note that you only have to do this once:
```{r eval=FALSE} 
if(!require(devtools)) install.packages("devtools")
```   
Then, load `devtools` and use the function `install_github()` to install `evalid`:
```{r eval=FALSE}
library(devtools)
install_github("naoki-egami/evalid", dependencies = TRUE)
```  

## Overview
Please read Section "The Proposed Approach toward External Validity: Outline" in Egami and Hartman (2022+) for the overview of effect- and sign-generalization.

**Effect-Generalization**
- Weighting-based Estimator: IPW estimator (`ipw`) or Weighted Least Squares (`wls`)
- Outcome-based Estimator: OLS-based estimator (`outcome-ols`) or BART-based estimator (`outcome-bart`)
- Doubly Robust Estimator: Doubly robust estimator with OLS-based outcome model (`dr-ols`) or Doubly robust estimator with BART-based outcome model (`dr-bart`)

**Sign-Generalization**
- How to use `pct()` function

## Effect-Generalization via `tpate()`
The goal is to estimate the target population average treatment effect (T-PATE) in the target population data. Please read Section "Effect-Generalization" in Egami and Hartman (2022+) for details and practical recommendations for effect-generalization. We use Broockman and Kalla (2018) as an example. See Section "Empirical Applications" in Egami and Hartman (2022+) as well.

### Prepare data
We need two data sets. (1) `exp_data`: the experimental data that includes outcome, treatment, and covariates we adjust for. (2) `pop_data`: the target population data that includes covariates we adjust for. 

```{r eval = TRUE, echo = TRUE, tidy=FALSE, warning=FALSE, error=FALSE, message=FALSE}
## Load the package and data
library(evalid)
data("BroockmanKalla")

exp_data <- BroockmanKalla$exp_data # experimental data
pop_data <- BroockmanKalla$pop_data # target population data

# covariates measured in both experimental and population data
covariates <- c("vf_female", "vf_black", "vf_white", 
                "religious_t0_factor_More_than_once_a_week", "religious_t0_factor_Never", 
                "religious_t0_factor_Once_a_week", "religious_t0_factor_Once_or_twice_a_month", 
                "ideology_t0_factor_Conservative", "ideology_t0_factor_Liberal",
                "ideology_t0_factor_Moderate", "ideology_t0_factor_Very_conservative", 
                "pid_t0_factor_Independent", "pid_t0_factor_Lean_Democrat", 
                "pid_t0_factor_Lean_Republican", "pid_t0_factor_Not_very_strong_Democrat",
                "pid_t0_factor_Not_very_strong_Republican", "pid_t0_factor_Strong_Democrat", 
                "vf_age_bucket_a_18to34", "vf_age_bucket_b_35to49", "vf_age_bucket_c_50to64")
```

### Weighting-based Estimators
Weighting-based estimators (`ipw` and `wls`) estimate the T-PATE via estimating weights that balance covariates in the experimental data and population data. In addition to the ignorability of sampling and treatment effect heterogeneity, this class of estimators assumes weights are correctly modeled. Importantly, weighted least squares estimators (`wls`) can incorporate covariates measured only in the experiment to increase efficiency (using argument `covariates_exp`). 
```{r eval = TRUE, echo = TRUE, tidy=FALSE, warning=FALSE, error=FALSE, message=FALSE}
## IPW 
ipw <- tpate(outcome = "trans.tolerance.dv.t1",
             treatment = "treat_ind", 
             covariates = covariates, 
             exp_data = exp_data, 
             pop_data = pop_data,
             id_cluster = exp_data$cluster, # cluster-standard errors 
             est_type = "ipw", # type of estimator 
             weights_max = 10) 

unlist(ipw$tpate)

## Weighted Least Squares (can include covariates measured only in the experiment)
covariates_exp <- c("miami_trans_law_t0", "miami_trans_law2_t0",
                    "therm_trans_t0", "gender_norms_sexchange_t0", "gender_norms_moral_t0", "gender_norms_abnormal_t0",
                    "ssm_t0", "therm_obama_t0", "therm_gay_t0", "vf_democrat",
                    "exposure_gay_t0", "exposure_trans_t0", "sdo_scale",
                    "gender_norm_daugher_t0", "gender_norm_looks_t0",
                    "gender_norm_rights_t0", "therm_afams_t0", "vf_hispanic",
                    "survey_language_es","cluster_level_t0_scale_mean")

wls <- tpate(outcome = "trans.tolerance.dv.t1",
             treatment = "treat_ind", 
             covariates = covariates, # covariates measured in both the experimental and population data
             covariates_exp = covariates_exp, # covariates measured only in the experimental data
             exp_data = exp_data, 
             pop_data = pop_data,
             id_cluster = exp_data$cluster, # cluster-standard errors 
             est_type = "wls", # type of estimator 
             weights_max = 10) 
unlist(wls$tpate)
```

### Outcome-based Estimators
Outcome-based estimators (`outcome-ols` and `outcome-bart`) estimate the T-PATE via modeling outcome regression. In addition to the ignorability of sampling and treatment effect heterogeneity, this class of estimators assumes outcome models are correctly specified.
```{r eval = TRUE, echo = TRUE, tidy=FALSE, warning=FALSE, error=FALSE, message=FALSE}
## OLS Outcome-based Estimator 
outcome_ols <- tpate(outcome = "trans.tolerance.dv.t1",
                     treatment = "treat_ind", 
                     covariates = covariates, # covariates measured in both the experimental and population data
                     exp_data = exp_data, 
                     pop_data = pop_data,
                     id_cluster = exp_data$cluster, # cluster-standard errors 
                     est_type = "outcome-ols") # type of estimator 
unlist(outcome_ols$tpate)

## BART Outcome-based Estimator 
outcome_bart <- tpate(outcome = "trans.tolerance.dv.t1",
                      treatment = "treat_ind", 
                      covariates = covariates, # covariates measured in both the experimental and population data
                      exp_data = exp_data, 
                      pop_data = pop_data,
                      id_cluster = exp_data$cluster, # cluster-standard errors 
                      est_type = "outcome-bart") # type of estimator
unlist(outcome_bart$tpate)
```

### Doubly Robust Estimators
Doubly robust estimators (`dr-ols` and `dr-bart`) estimate the T-PATE via modeling both weights and outcome regression. Doubly robust estimators are consistent to the T-PATE if one of the two (weights or an outcome model) is correctly specified. Note that this estimator also makes the ignorability of sampling and treatment effect heterogeneity assumption for causal identification.
```{r eval = TRUE, echo = TRUE, tidy=FALSE, warning=FALSE, error=FALSE, message=FALSE}
## Doubly Robust Estimator (outcome model is OLS)
dr_ols <- tpate(outcome = "trans.tolerance.dv.t1",
                treatment = "treat_ind", 
                covariates = covariates, 
                exp_data = exp_data, 
                pop_data = pop_data,
                id_cluster = exp_data$cluster, # cluster-standard errors 
                est_type = "dr-ols", # type of estimator 
                weights_max = 10) 
unlist(dr_ols$tpate)

## Doubly Robust Estimator (outcome model is BART)
dr_bart <- tpate(outcome = "trans.tolerance.dv.t1",
                 treatment = "treat_ind", 
                 covariates = covariates, 
                 exp_data = exp_data, 
                 pop_data = pop_data,
                 id_cluster = exp_data$cluster, # cluster-standard errors 
                 est_type = "dr-bart", # type of estimator 
                 weights_max = 10) 
unlist(dr_bart$tpate)
```

## Sign-Generalization via `pct()`
The goal is to test the sign of the target population average treatment effect (T-PATE) using design of purposive variations within an experiment. Please read Section "Sign-Generalization" in Egami and Hartman (2022+) for details and practical recommendations for sign-generalization. We use Bisgaard (2019) as an example. See Section "Empirical Applications" in Egami and Hartman (2022+) as well.

### Prepare data
We need to compute the SATE for each purposive variation. To facilitate this process, we created a cleaned data set for each purposive variation (in total, 12 for opposition supporters and government supporters, respectively).
```{r eval = TRUE, echo = TRUE, tidy=FALSE, warning=FALSE, error=FALSE, message=FALSE}
data("Bisgaard")

## Opposition Supporters (The Bisgaard theory predicts the positive T-PATE for this subgroup)
opp_data <- Bisgaard$opp_data  ## cleaned data for 12 different variations 

head(opp_data[[1]][, c("outcome", "treatment")])

## Government Supporters (The Bisgaard theory predicts the negative T-PATE for this subgroup)
gov_data <- Bisgaard$gov_data  ## cleaned data for 12 different variations 

head(gov_data[[1]][, c("outcome", "treatment")])
```


### Testing whether the T-PATE is positive
First, we consider cases when a substantive theory predicts the T-PATE is positive. Therefore, a null hypothesis is that the T-PATE is less than or equal to zero. Therefore, when we compute one-sided p-values, we use `1 - prnom(z)` where `z` is the ratio of an estimated causal effect and its standard error. 
```{r eval = TRUE, echo = TRUE, tidy=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(estimatr)
## Compute one-sided p-values for each variation
pv_opp <- c()
for(i in 1:12){
  data_variation <- opp_data[[i]]
  fit_variation <- lm_robust(outcome ~ treatment, data = data_variation)  # compute the SATE for each variation
  z <- summary(fit_variation)$coef["treatment", "Estimate"]/summary(fit_variation)$coef["treatment", "Std. Error"]
  pv_opp[i] <- 1 - pnorm(z)  # one-sided p-value for positive effects
}

## Partial Conjunction Test
round(pct(pv_opp), 2)
```

### Testing whether the T-PATE is negative
Next, we consider cases when a substantive theory predicts the T-PATE is negative. Therefore, a null hypothesis is that the T-PATE is larger than or equal to zero. Therefore, when we compute one-sided p-values, we use `prnom(z)` where `z` is the ratio of an estimated causal effect and its standard error. 
```{r eval = TRUE, echo = TRUE, tidy=FALSE, warning=FALSE, error=FALSE, message=FALSE}
## Compute one-sided p-values for each variation
pv_gov <- c()
for(i in 1:12){
  data_variation <- gov_data[[i]]
  fit_variation <- lm_robust(outcome ~ treatment, data = data_variation)  # compute the SATE for each variation
  z <- summary(fit_variation)$coef["treatment", "Estimate"]/summary(fit_variation)$coef["treatment", "Std. Error"]
  pv_gov[i] <- pnorm(z)  # one-sided p-value for negative effects
}

## Partial Conjunction Test
round(pct(pv_gov), 2)
```
